Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-03-17 doc/paper/paper.tex
Mostly writing

\usepackage{alltt}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{multirow}
\usepackage{cleveref}

@@

When an infectious disease spreads through a population, transmissions cannot
generally occur between any two individuals at any time. Rather, the
individuals must have a particular mode of contact with one another, which
varies with the disease's mode of transmission. For airborne pathogens,
physical proximity may be sufficient, but for sexually transmitted diseases,
either sexual or blood-to-blood contact is required. The set of links between
individuals along which transmission can occur is called the contact
network~\autocite{klovdahl1985social}. The structure of the contact network
underlying an epidemic can profoundly impact the speed and pattern of the
epidemic's expansion. Network structure can influence the prevalence
trajectory~\autocite{o2010contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
of quantities such as effective population
size~\autocite{goodreau2006assessing}. From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{wang2015targeting}. True contact networks are a challenging
type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

Viral sequence data, on the other hand, has become relatively inexpensive and
straightforward to collect on a population level. Due to the high mutation rate
of RNA viruses, epidemiological processes impact the course of viral evolution,
thereby shaping the intra-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2011estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, inference
of structural network parameters has to date been limited. However, it has been
shown that network structure has a tangible impact on phylogeny
shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
goodreau2006assessing, robinson2013dynamics}, suggesting that such statistical
inference might be possible~\autocite{welch2011statistical}.

@@

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a
Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{brown2011transmission} found that a Waring distribution,
which is produced by a different preferential attachment model, was a good fit
to a phylogenetically estimated transmission network relating 60\% of HIV
infected \gls{MSM} in the United Kingdom. The transmission network is a
subgraph of the contact network which includes only those edges which have
already led to a new infection.

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using \gls{MCMC}, which requires the calculation of likelihood ratios.
Unfortunately, it is generally difficult to explicitly calculate the likelihood
of an observed transmission tree under a contact network model. To do so, it
would be necessary to integrate over all possible networks, and also over all
possible labellings of the internal nodes of the transmission tree. While it is
not known (to us) whether such integration is tractable, a simpler alternative
is offered by likelihood-free methods, namely \gls{ABC}. \gls{ABC} leverages
the fact that, although calculating the likelihood may be impossible,
generating simulated datasets according to a model is often straightforward. If
our model fits the data well, the simulated data it produces should be similar
to the observed data. More formally, if $D$ is the observed data, the posterior
distribution $f(\theta \mid D)$ on model parameters $\theta$ is replaced as the
target of statistical inference by $f(\theta \mid \rho(\hat{D}, D) <
\varepsilon)$, where $\rho$ is a distance function, $\hat{D}$ is a simulated
dataset according to $\theta$, and $\varepsilon$ is a small
tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
is a kernel function, the approach is known as kernel-\gls{ABC}.

@@

We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
for simulating epidemics and transmission trees over static contact networks,
in the same fashion as several previous studies
\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
leventhal2012inferring, groendyke2011bayesian, goodreau2006assessing}. To check
that our implementation was correct, we reproduced Figure 1A of
\textcite{leventhal2012inferring} (our \cref{fig:sf1}), which plots the
unbalancedness of transmission trees simulated over four network models at
various levels of pathogen transmissibility.

@@

We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
whether the parameters of the \gls{BA} model had an effect on tree shape. The
parameters were varied one at a time over the following values: $N$ =
\sett{3000, 5000, 8000}, $I$ = \sett{500, 1000, 2000}, $m$ = \sett{2, 3, 4},
and $\alpha$ = \sett{0.5, 1, 1.5}. The parameters not being tested were fixed
at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ = 1. For each
parameter set, 100 networks were generated, and a transmission tree was
simulated over each (300 trees per parameter set). The 300 trees were compared
pairwise with the tree kernel to form a $300 \times 300$ kernel matrix. We
constructed a \gls{kSVM} classifier for the parameter of interest using the
\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
accuracy with 1000 two-fold cross-validations. The construction of the kernel
matrices and classifiers was repeated for several combinations of the kernel
meta-parameters $\lambda$ (the ``decay factor''), and $\sigma$ (the ``radial
basis function variance'')~\autocite[see][]{poon2013mapping}, and for
transmission trees of sizes 100, 500, and 1000. We also tested univariate
classifier based on Sackin's index~\autocite{shao1990tree} and an ordinary SVM
based on the \gls{nLTT} statistic~\autocite{janzen2015approximate}.

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested~\cref{fig:rsquared}. Classifiers based on two other tree statistics, the
\gls{nLTT} and Sackin's index, generally exhibited worse performance than the
tree kernel, although the magnitude of the disparity varied between the
parameters (\cref{fig:crossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). For
appropriate choices of the kernel meta-parameters, the kSVM classifier for
$\alpha$ had an $R^2$ above 
    0.8 
for all combinations of tree size and epidemic size (\cref{fig:crossv}). The
other classifiers' $R^2$ averaged between
  0.49
and
  0.61
for the SVM based on the \gls{nLTT}, and between
  0.73
and
  0.76
for the linear regression against Sackin's index. The $I$ parameter was also
classified accurately by the tree kernel (\cref{fig:Icrossv}), with an average
$R^2$ of 
  0.93 
for 500-tip trees or 
  0.7 
for 100-tip trees. The \gls{nLTT} performed almost as well on this parameter as
the tree kernel, with $R^2$ values of
  0.83
for 500-tip trees or
  0.55
for 100-tip trees. Sackin's index was an extremely poor classifier for this
parameter, with both $R^2$ values below
  0.08.

The $m$ parameter was much harder to classify (\cref{fig:mcrossv}), with $R^2$
for the kSVM varying between 
  0.004
and
  0.36
depending on $I$ and the tree size. The \gls{nLTT} was also universally poor,
with $R^2$ below
  0.02
for all scenarios. Sackin's index fared slightly better, especially on trees
with a higher number of tips, with $R^2$ between
  0
and
  0.18.
Finally, the accuracy of the kSVM classifier for $N$ varied widely
(\cref{fig:Ncrossv}), from 
  0.08
to
  0.82.
The \gls{nLTT} performed well in classifying $N$ when the epidemic was large,
improving upon the tree kernel in the scenarios with $I$ = 2000 and 100- or
500-tip transmission trees ($R^2$ = 
  0.31
and
  0.53),
although it performed poorly when $I$ was 500 or 1000 (average $R^2$ =
  0.07).
Sackin's index could not classify $N$, with all $R^2$ below
  0.11.
Based on inspection of the cross-validation results, we chose to use the
meta-parameters $\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not
to use Sackin's index or the \gls{nLTT}.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\figwidth]{kernel-rsquared}
  \caption{
    Cross-validation accuracy of kernel-SVM classifiers for \gls{BA} model
    parameters, with kernel meta-parameters $\lambda = 0.3$ and $\sigma = 4$.
    Each point was calculated based on 300 simulated transmission trees over
    networks with three different values of the parameter being tested.
    Vertical lines are empirical 95\% confidence intervals based on 1000 2-fold
    cross-validations.
  }
  \label{fig:rsquared}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\figwidth]{kernel-alpha-crossv}
  \caption{
    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
    of \gls{BA} network model, for various tree kernel meta-parameters and
    epidemic scenarios. Each point was calculated based on 300 simulated
    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
    and and dashed lines indicate, respectively, performance of SVM using the
    \gls{nLTT} statistic, and linear regression using Sackin's index. Facets
    are number of infected nodes before the simulation was stopped ($I$) and
    number of tips in the sampled transmission tree.
  }
  \label{fig:crossv}
\end{figure}

@@

Of the four parameters, $\alpha$ was the most accurately estimated, with a
median [IQR] absolute error of 
    0.11 
    [0.05-0.18].
(\cref{fig:abcpt,tab:abchpd}). The accuracy of the estimates was not
significantly different between values of $m$ or $I$ (both one-way ANOVA,
    $p$ = 0.1
and 
    0.25),
although the errors when the true value of $\alpha$ was zero were
significantly greater than the other values 
    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.41\times 10^{-4}}).
The error in the estimated value of $I$ was
    305.66 
    [107.76-606.59].
Errors were significantly higher for $\alpha \geq 1$
    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.12\times 10^{-4}})
and for $I$ = 2000
    ($p$ = \ensuremath{1.58\times 10^{-6}}),
but not for any values of $m$
    (one-way ANOVA, $p$ = 0.33).
The $m$ parameter was estimated correctly in
    37 \%
of simulations, with an error of one in
    40 \%
and of two or more in 
    22 \%
(the only possible $m$ values were 1, 2, 3, 4, or 5). The true values of $m$
and $I$ did not significantly affect the error
    (one-way ANOVA, $p$ = 0.5 and
                          0.68).
%but the accuracy was significantly lower for integral than non-integral
%values of $\alpha$
%    (Wilcoxon rank-sum test, $p$ = alpha.m.test$p.value).
Finally, the total number of nodes $N$ was consistently over-estimated by about
a factor of two
    (error \ensuremath{6.59\times 10^{3}} 
    [\ensuremath{4.21\times 10^{3}}-\ensuremath{8.28\times 10^{3}}]).
No other parameters influenced the accuracy of the $N$ estimates 
    (one-way ANOVA, $p \geq$ 0.72).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\figwidth]{abc-point-estimate}
  \caption{
    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
    on simulated phylogenies. Dotted lines indicate the true values, and limits
    of the $y$-axes are the regions of uniform prior density.
  }
  \label{fig:abcpt}
\end{figure}

@@

The dispersion of the \gls{ABC} approximation to the posterior also varied
between the parameters, with narrower \gls{HPD} intervals for those with the
most accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows the
distributions for for one simulation (equivalent plots for all the simulations
can be found in the supplemental materials). \gls{HPD} intervals around
$\alpha$ and $I$ were narrow relative to the region of nonzero prior density,
whereas the intervals for $m$ and $N$ were more widely dispersed.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\figwidth]{abc-posterior-example}
  \caption{Marginal estimated posterior distributions obtained with ABC on a
  single simulated dataset. Dotted lines and shaded polygon indicate true
  values.}
  \label{fig:abcex}
\end{figure}

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The \gls{MAP} [95\%
HPD] estimates for each parameter were: 
$\alpha$, 
  1 [0.85-
                           1.12];
$I$,
  5.23 [2.05-
                       5.85];
$m$,
  \ensuremath{10^{4}} [\ensuremath{3.06\times 10^{3}}-
                              \ensuremath{1.5\times 10^{4}}];
$N$,
  \ensuremath{1.17\times 10^{3}} [505.71-
                       \ensuremath{1.75\times 10^{3}}].

To test the effect of sampling bias, we sampled one transmission tree in a
peer-driven fashion, where the probability to sample a node was twice as high
if one of it's peers had already been sampled. The parameters for this
experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
estimated values were
$\alpha$, 
  0.48 [0.03-
                           0.83];
$I$,
  3.6 [2.18-
                       5.9];
$m$,
  \ensuremath{1.08\times 10^{4}} [\ensuremath{3.11\times 10^{3}}-
                              \ensuremath{1.48\times 10^{4}}];
$N$,
  \ensuremath{2.48\times 10^{3}} [\ensuremath{1.28\times 10^{3}}-
                       \ensuremath{3.87\times 10^{3}}].

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\figwidth]{realdata-hpd}
  \caption{
    Point estimates and 95\% \gls{HPD} intervals for \gls{BA} network
    parameters of published HIV datasets, obtained with kernel-ABC.
  }
\end{figure}

@@

Our investigation of published HIV datasets indicated heterogeneity in the
contact network structures underlying several distinct local epidemics.
\textcite{cuevas2009hiv} studied a group of newly diagnosed individuals in the
Basque Country, Spain. Although the individuals were of mixed risk group, and
therefore unlikely to comprise a single contact network, a high proportion of
them (47\%) grouped into local transmission clusters, allowing for the
possibility of weak preferential attachment dynamics. We estimated a relatively
low attachment power for these data, which is consistent with the sampled
sequences comprising many distinct sub-networks. \textcite{li2015} sampled a
large number of acutely infected \gls{MSM} in Shanghai, China. By using a
patristic distance cutoff~\autocite{poon2014impact}, we identified and analysed
a large cluster. Surprisingly, the estimated attachment power was fairly low
with a large credible interval.

@@

Our method itself has of caveats, perhaps the most significant being that it
takes a transmission tree as input. In reality, true transmission trees are not
available and must be approximated, often by way of a viral phylogeny. Although
this has been demonstrated to be a fair approximation in
practice~\autocite[e.g.][]{leitner1996accurate}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating}. In addition, it is computationally
intensive, taking about a day when run on 20 cores in parallel with the setings
we described in the methods. However, there are no other methods available to
perform the types of estimation we do here, and we therefore believe our method
will be useful to epidemiological researchers interested in the network
structure underlying disease outbreaks.

@@

\end{document}
\newpage
\section*{Supplemental Materials}

\begin{figure}[ht]
  \centering
  %\includegraphics[scale=0.8]{leventhal2012fig1}
  \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
  \label{fig:sf1}
\end{figure}

>>>>

\usepackage{alltt}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{multirow}
\usepackage{cleveref}
\usepackage{array}

@@

When an infectious disease spreads through a population, transmissions are
generally more likely to occur between certain pairs of individuals. Such pairs
must have a particular mode of contact with one another, which varies with the
mode of transmission of the disease. For airborne pathogens, physical proximity
may be sufficient, but for sexually transmitted diseases, either sexual or
blood-to-blood contact is required. The set of links between individuals along
which transmission can occur is called the contact
network~\autocite{klovdahl1985social, morris1993epidemiology}. The structure of
the contact network underlying an epidemic can profoundly impact the speed and
pattern of the epidemic's expansion. Network structure can influence the
prevalence trajectory~\autocite{o2010contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
of quantities such as effective population
size~\autocite{goodreau2006assessing}. From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{wang2015targeting}. True contact networks are a challenging
type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

Viral sequence data, on the other hand, has become relatively inexpensive and
straightforward to collect on a population level. Due to the high mutation rate
of RNA viruses, epidemiological processes impact the course of viral evolution,
thereby shaping the intra-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2011estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, although
sophisticated methods have been developed for fitting complex population
genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
of structural network parameters has to date been limited. However, it has been
shown that network structure has a tangible impact on phylogeny
shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
goodreau2006assessing, robinson2013dynamics}, suggesting that such statistical
inference might be possible~\autocite{welch2011statistical}.

@@

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a
Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{brown2011transmission} analysed the degree distribution
of an approximate transmission network, estimated based on genetic similarity
and estimated times of infection, relating 60\% of HIV infected \gls{MSM} in
the United Kingdom. The transmission network is a subgraph of the contact
network which includes only those edges which have already led to a new
infection. The authors found that a Waring distribution, which is produced by a
different preferential attachment model, was a good fit to their estimated
network.

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using \gls{MCMC}. Rather than calculating explicit likeilihoods, \gls{MCMC}
uses likelihood \emph{ratios} in which the normalizing constant cancel out.
Unfortunately, it is generally difficult to explicitly calculate the likelihood
of an observed transmission tree under a contact network model, even up to a
normalizing constant. To do so, it would be necessary to integrate over all
possible networks, and also over all possible labellings of the internal nodes
of the transmission tree. While it is not known (to us) whether such
integration is tractable, a simpler alternative is offered by likelihood-free
methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
calculating the likelihood may be impossible, generating simulated datasets
according to a model is often straightforward. If our model fits the data well,
the simulated data it produces should be similar to the observed data. More
formally, if $D$ is the observed data, the posterior distribution $f(\theta
\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
specific case when $\rho$ is a kernel function, the approach is known as
kernel-\gls{ABC}~\autocite{nakagome2013kernel, poon2015phylodynamic}.

@@

We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
for simulating epidemics and transmission trees over static contact networks,
in the same fashion as several previous studies
\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
leventhal2012inferring, groendyke2011bayesian, goodreau2006assessing}. To check
that our implementation was correct, we reproduced Figure 1A of
\textcite{leventhal2012inferring} (our \cref{fig:sf1}), which plots the
unbalancedness of transmission trees simulated over four network models at
various levels of pathogen transmissibility. Our program is freely available at
\url{https://github.com/rmcclosk/netabc}.

@@

We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
whether the parameters of the \gls{BA} model had an effect on tree shape. We
simulated 100 networks under each of three different values of $\alpha$: 0.5,
1.0, and 1.5 (300 networks total). The other parameters were fixed to the
following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
500 tips was simulated over each network (300 transmission trees total). The
300 trees were compared pairwise with the tree kernel to form a $300 \times
300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
factor''), and $\sigma$ (the ``radial basis function
variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4
respectively. We constructed a \gls{kSVM} classifier for $\alpha$ using the
\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
accuracy with 1000 two-fold cross-validations.

Three similar experiments were performed for the other \gls{BA} model
parameters (one experiment per parameter). $m$ was varied between 2, 3, and 4;
$I$ between 500, 1000, and 2000; and $N$ between 3000, 5000, and 8000. The
parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
$m$ = 2, and $\alpha$ = 1. Thus, we performed a total of four kSVM
cross-validations, one for each of the \gls{BA} model parameters $\alpha$, $I$,
$m$, and $N$.

We repeated these four cross-validations with different values of $\lambda$
(0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well
as on trees with differing numbers of tips (100, 500, and 1000) and in
epidemics of differing size $I$ (500, 1000, and 2000). When evaluating the
classifier for $I$, we did not consider trees with 1000 tips, because one of
the tested $I$ values was 500, and the number of tips cannot be larger than
$I$.

For each of the four parameters, we also tested univariate classifier based on
Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the
\gls{nLTT} statistic~\autocite{janzen2015approximate}.

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested~\cref{fig:rsquared}. Classifiers based on two other tree statistics, the
\gls{nLTT} and Sackin's index, generally exhibited worse performance than the
tree kernel, although the magnitude of the disparity varied between the
parameters. \Cref{fig:rsquared} shows the cross-validation accuracy with
$\lambda$ = 0.3, $\sigma$ = 4, $I$ = 1000, and trees of size 500. The results
were largely robust to variations in $\lambda$ and $\sigma$, although accuracy
varied between different epidemic and sampling scenarios
(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).

The kSVM classifier for $\alpha$ had an average $R^2$ of
    0.92 
(\cref{fig:alphacrossv}). The $R^2$ of the for the SVM classifer for $\alpha$
using the \gls{nLTT} varied between
0.49
and
  0.61.
The $R^2$ of the linear regression against Sackin's index varied between
  0.73
and
  0.76.
The $I$ parameter was also classified accurately by the tree kernel
(\cref{fig:Icrossv}), with an average
$R^2$ of 
  0.93 
for 500-tip trees or 
  0.7 
for 100-tip trees. The \gls{nLTT} performed almost as well on this parameter as
the tree kernel, with $R^2$ values of
  0.83
for 500-tip trees or
  0.55
for 100-tip trees. Sackin's index was an extremely poor classifier for this
parameter, with both $R^2$ values below
  0.08.

The $m$ parameter was much harder to classify (\cref{fig:mcrossv}), with $R^2$
for the kSVM varying between 
  0.004
and
  0.36
depending on $I$ and the tree size. The \gls{nLTT} was also universally poor,
with $R^2$ below
  0.02
for all scenarios. Sackin's index fared slightly better than the \gls{nLTT},
especially on trees with a higher number of tips, with $R^2$ between
  0
and
  0.18.
Finally, the accuracy of the kSVM classifier for $N$ varied widely
(\cref{fig:Ncrossv}), from 
  0.08
to
  0.82.
The \gls{nLTT} performed well in classifying $N$ when the epidemic was large,
improving upon the tree kernel in the scenarios with $I$ = 2000 and 100- or
500-tip transmission trees ($R^2$ = 
  0.31
and
  0.53),
although it performed poorly when $I$ was 500 or 1000 (average $R^2$ =
  0.07).
Sackin's index could not classify $N$, with all $R^2$ below
  0.11.
Based on inspection of the cross-validation results, we chose to use the
meta-parameters $\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not
to use Sackin's index or the \gls{nLTT}.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
  \caption{
      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
      using \gls{nLTT} (centre), and linear regression using Sackin's index
      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
      simulated transmission trees over networks with three different values of
      the parameter being tested. Vertical lines are empirical 95\% confidence
      intervals based on 1000 2-fold cross-validations.
  }
  \label{fig:rsquared}
\end{figure}

@@

Of the four parameters, $\alpha$ was the most accurately estimated, with a
median [IQR] absolute error of 
    0.11 
    [0.05-0.18].
\Cref{fig:abcpt} shows point estimates for all simulations with $m = 2$. The
results $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}).
Average boundaries of 95\% \gls{HPD} intervals are given in \cref{tab:abchpd}.
The accuracy of the estimates was not significantly different between values of
$m$ or $I$ (both one-way ANOVA,
    $p$ = 0.1
and 
    0.25),
although the errors when the true value of $\alpha$ was zero were
significantly greater than the other values 
    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.41\times 10^{-4}}).
The error in the estimated value of $I$ was
    305.66 
    [107.76-606.59].
Errors were significantly higher for $\alpha \geq 1$
    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.12\times 10^{-4}})
and for $I$ = 2000
    ($p$ = \ensuremath{1.58\times 10^{-6}}),
but not for any values of $m$
    (one-way ANOVA, $p$ = 0.33).
The $m$ parameter was estimated correctly in
    37 \%
of simulations, with an error of one in
    40 \%
and of two or more in 
    22 \%
(the only possible $m$ values were 1, 2, 3, 4, or 5). The true values of $m$
and $I$ did not significantly affect the error
    (one-way ANOVA, $p$ = 0.5 and
                          0.68).
Finally, the total number of nodes $N$ was consistently over-estimated by about
a factor of two
    (error \ensuremath{6.59\times 10^{3}} 
    [\ensuremath{4.21\times 10^{3}}-\ensuremath{8.28\times 10^{3}}]).
No other parameters influenced the accuracy of the $N$ estimates 
    (one-way ANOVA, $p \geq$ 0.72).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{abc-point-estimate-m2.pdf}
  \caption{
    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
    on simulated phylogenies without training, for simulations with $m = 2$.
    Dotted lines indicate true values, and limits of the $y$-axes are regions
    of uniform prior density.
  }
  \label{fig:abcpt}
\end{figure}

@@

The dispersion of the \gls{ABC} approximation to the posterior also varied
between the parameters, with narrower \gls{HPD} intervals for those with the
most accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows the
distributions for for one simulation (equivalent plots for all the simulations
can be found in supplemental data). \Gls{HPD} intervals around $\alpha$ and $I$
were narrow relative to the region of nonzero prior density, whereas the
intervals for $m$ and $N$ were more widely dispersed.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{abc-posterior-example.pdf}
  \caption{Marginal estimated posterior distributions obtained with ABC on a
  single simulated dataset. Dotted lines and shaded polygon indicate true
  values.}
  \label{fig:abcex}
\end{figure}

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The \gls{MAP} [95\%
HPD] estimates for each parameter were: 
$\alpha$, 
  1 [0.07-
                           1.07];
$I$,
  5.37 [2.05-
                       5.82];
$m$,
  \ensuremath{1.01\times 10^{4}} [\ensuremath{3.06\times 10^{3}}-
                              \ensuremath{1.5\times 10^{4}}];
$N$,
  \ensuremath{1.18\times 10^{3}} [505.71-
                       \ensuremath{1.62\times 10^{3}}].

To test the effect of sampling bias, we sampled one transmission tree in a
peer-driven fashion, where the probability to sample a node was twice as high
if one of it's peers had already been sampled. The parameters for this
experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
estimated values were
$\alpha$, 
  0.46 [0.03-
                           0.83];
$I$,
  3.53 [2.18-
                       5.9];
$m$,
  \ensuremath{1.07\times 10^{4}} [\ensuremath{3.11\times 10^{3}}-
                              \ensuremath{1.47\times 10^{4}}];
$N$,
  \ensuremath{2.46\times 10^{3}} [\ensuremath{1.28\times 10^{3}}-
                       \ensuremath{3.88\times 10^{3}}].

@@

\begin{figure}[ht]
  \centering
  \includegraphics{realdata-hpd.pdf}
  \caption{
    Point estimates and 95\% \gls{HPD} intervals for \gls{BA} network
    parameters of published HIV datasets, obtained with kernel-ABC.
  }
\end{figure}

@@

As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
parameters may translate to highly informative priors on quantities of
interest. We observed an exponential relationship between the preferential
attachment power $\alpha$ and the power-law exponent $\gamma$
(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
and 2 is equivalent to placing an informative prior that $\gamma$ is close to
2. Therefore, if we were primarily interested in $\gamma$ rather than $\alpha$,
a more sensible choice of prior might have a shape similar to \cref{fig:gamma}
and be bounded above by approximately $\alpha$ = 1.5. This would uniformly
bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly reported in the
network literature~\autocite{liljeros2001web, schneeberger2004scale,
colgate1989risk, brown2011transmission}. We note however that
\textcite{jones2003assessment} estimated $\gamma$ values greater than four, in
one case as high as 17, for some datasets, indicating that a wider range of
permitted $\gamma$ values may be warranted.

Our investigation of published HIV datasets indicated heterogeneity in the
contact network structures underlying several distinct local epidemics.
\textcite{cuevas2009hiv} studied a group of newly diagnosed individuals in the
Basque Country, Spain. Although the individuals were of mixed risk group, and
therefore unlikely to comprise a single contact network, a high proportion of
them (47\%) grouped into local transmission clusters, allowing for the
possibility of weak preferential attachment dynamics. We estimated a relatively
low attachment power for these data, which is consistent with the sampled
sequences comprising many distinct sub-networks. \textcite{li2015hiv} sampled a
large number of acutely infected \gls{MSM} in Shanghai, China. By using a
patristic distance cutoff~\autocite{poon2014impact}, we identified and analysed
a large cluster. Surprisingly, the estimated attachment power was fairly low
with a large credible interval.

@@

Our method itself has of caveats, perhaps the most significant being that it
takes a transmission tree as input. In reality, true transmission trees are not
available and must be approximated, often by way of a viral phylogeny. Although
this has been demonstrated to be a fair
approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating}. In addition, it is computationally
intensive, taking about a day when run on 20 cores in parallel with the setings
we described in the methods. However, there are no other methods available to
perform the types of estimation we do here, and we therefore believe our method
will be useful to epidemiological researchers interested in the network
structure underlying disease outbreaks.

@@

\newpage
\section*{Supplemental Materials}

\begin{figure}[ht]
  \centering
  \includegraphics{leventhal2012fig1.pdf}
  \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
  \label{fig:sf1}
\end{figure}
